### 2015-07-30 doc/thesis/system.tex
More fixes to the text about typing rules

With recursive types we finish the dicussion about Typed Lua types,
and we begin the discussion about special types.

@@

The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
It allows covariant keys because we also want to use \emph{unique}
table types as a way to join table fields that inhabit \emph{closed} table types.
For instance, we want to use the table constructor to initialize
a variable with a table type that describes a hash.
More precisely, this rule states that it is safe to recast the table type
$\{``x":\Integer, ``y":\Integer, ``z":\Integer\}_{unique}$ to
$\{\String:\Integer \cup \Nil\}_{closed}$, as long as the
new type becomes inacessible with the original type.

@@

We do not need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, because \textsc{T-IDREAD}
is sufficient to forbid the creation of an alias to another \emph{unique}
or \emph{open} table.
For this reason, identifiers that appear in the left-hand side
of assignements have their own rule \textsc{T-IDWRITE1}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE1}\\
\dfrac{\env_{1}(id) = F}
      {\env_{1}, \penv \vdash id_{l}:F, \env_{1}}
\end{array}
\]

@@

This rule uses the auxiliary function \emph{infer} to get the
most general first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.

@@

Now, we may want to discriminate $q$ and $r$ to check whether
the function call returned with success.
Introducing a projection variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$,
as they are a general way to not compromise the depency between
the types of $q$ and $r$ after discriminating one of them,
so flow typing can narrow the type of both variables by testing
just one of them because the projection types of both variables
bound to the same projection variable.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x}\\
       S_{t} = fonilpt(\penv(x), i) \;\;\;
       S_{e} = finilpt(\penv(x), i) \\
       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2} \;\;\;
       \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{3}\\
       \env_{4} = join(\env_{2}, \env_{3})
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

Rule \textsc{T-IF3} uses the auxiliary functions \emph{fonilpt} and \emph{finilpt}
to filter a projection $x$, affecting all variables that bind to the same projection.
The former function filters out the tuples that contain $\Nil$ in the $i$-th component,
while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
For instance, our previous example type checks through rule \textsc{T-IF3},
because it uses the information provided by the projection type $\pi_{1}^{x}$,
which is the type of $q$, to make the rule \textsc{T-IF3} use the function call
\[
fonilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
and the function call
\[
finilpt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.

@@

In this example, the projected type of $r$ outsided of the $\mathbf{if}$
statement is $\Integer \cup \String$, so the assignment looks fine.
However, the projected type of $r$ inside the $\mathbf{if}$ branch is $\Integer$,
not matching the $\String$ value that $r$ has after the assignment.

>>>>

With recursive types we finish the discussion about Typed Lua types,
and we begin the discussion about special types.

@@

The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
It allows covariant keys because we also want to use \emph{unique}
table types as a way to join table fields that inhabit \emph{closed} table types.
For instance, we want to use the table constructor to initialize
a variable with a table type that describes a hash.
More precisely, this rule states that it is safe to recast the table type
$\{``x":\Integer, ``y":\Integer, ``z":\Integer\}_{unique}$ to
$\{\String:\Integer \cup \Nil\}_{closed}$, as long as the
new type becomes inaccessible with the original type.

@@

We do not need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, because \textsc{T-IDREAD1}
and \textsc{T-IDREAD2} are sufficient to forbid the creation of an alias
to another \emph{unique} or \emph{open} table.
For this reason, identifiers that appear in the left-hand side
of assignments have their own rule \textsc{T-IDWRITE1}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE1}\\
\dfrac{\env_{1}(id) = F}
      {\env_{1}, \penv \vdash id_{l}:F, \env_{1}}
\end{array}
\]

@@

This rule uses the auxiliary function \emph{infer} to get the
most general types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.
We can define \emph{infer} as follows:
\begin{align*}
infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
\begin{array}{ll}
general(T_{i}) & \text{if $i < n$}\\
general(nil(T_{n})) & \text{if $i >= n$}
\end{array} \right.
\end{align*}

\begin{align*}
general(\False) & = \Boolean\\
general(\True) & = \Boolean\\
general({\it int}) & = \Integer\\
general({\it float}) & = \Number\\
general({\it string}) & = \String\\
general(F_{1} \cup F_{2}) & = general(F_{1}) \cup general(F_{2})\\
general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
general(\{F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}\}_{tag}) & = \{F_{1}{:}general(V_{1}), ..., F_{n}{:}general(V_{n})\}_{tag}\\
general(\mu x.F) & = \mu x.general(F)\\
general(T) & = T\\
\\
general2(F{*}) & = general(F){*}\\
general2(F \times P) & = general(F) \times general2(P)\\
general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
\end{align*}

@@

Now, we may want to discriminate $q$ and $r$ to check whether
the function call returned with success.
Introducing a projection variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$,
as they are a general way to not compromise the dependency between
the types of $q$ and $r$ after discriminating one of them,
so flow typing can narrow the type of both variables by testing
just one of them because the projection types of both variables
bound to the same projection variable.

The rule \textsc{T-IF5} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}.
It uses the auxiliary functions \emph{fopt} and \emph{fipt}
to filter a projection $x$, affecting all variables that bind to the same projection.
More precisely, the former function filters out the tuples that contain a type $\Nil$ in the $i$-th component,
while the latter function filters out the tuples that do not contain $\Nil$ in the $i$-th component.
We define \textsc{T-IF5} as follows:
\[
\begin{array}{c}
\mylabel{T-IF5}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \\
       S_{t} = fopt(\penv(x), \Nil, i) \;\;\;
       S_{e} = fipt(\penv(x), \Nil, i) \\
       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2}\\
       \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{3}\\
       \env_{4} = join(\env_{2}, \env_{3})
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

As an example, our previous example type checks through rule \textsc{T-IF5},
because it uses the information provided by the projection type $\pi_{1}^{x}$,
which is the type of $q$, to make the rule \textsc{T-IF5} use the function call
\[
fopt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Integer \times \Integer \times \Nil{*}$ inside the $\mathbf{if}$ branch,
and the function call
\[
fipt((\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*}), \Nil, 1)
\]
to discriminate the projection $x$ to the single tuple
$\Nil \times \String \times \Nil{*}$ inside the $\mathbf{else}$ branch.
Thus, reading $q$ and $r$ projects $\pi_{1}^{x}$ to $\Integer$ and
$\pi_{2}^{x}$ to $\Integer$ inside the $\mathbf{if}$ branch,
but it projects $\pi_{1}^{x}$ to $\Nil$ and $\pi_{2}^{x}$ to $\String$
inside the $\mathbf{else}$ branch.
Outside the condition, $q$ and $r$ use the original projection, that is,
they project to $\Integer \cup \Nil$ and $\Integer \cup \String$, respectively.

Our type system also includes rules that check whether a branch is unreachable.
Rules \textsc{T-IF6} and \textsc{T-IF7} respectively cover the case where the
$\mathbf{else}$ branch is unreachable and the case where the $\mathbf{then}$
branch is unreachable, because either the projected type of $\pi_{i}^{x}$ is
not a supertype of $\Nil$ or it is $\Nil$.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IF6}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \\
       S_{t} = fopt(\penv(x), \Nil, i) \\
       fit(proj(\penv(x), i), \Nil) = \Void \\
       \env_{1}, \penv[x \mapsto S_{t}] \vdash s_{1}, \env_{2}
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IF7}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \\
       S_{e} = fipt(\penv(x), \Nil, i) \\
       fot(proj(\penv(x), i), \Nil) = \Void \\
       \env_{1}, \penv[x \mapsto S_{e}] \vdash s_{2}, \env_{2}
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{2}}
\end{array}
\end{array}
\]

@@

In this example, the projected type of $r$ outside of the $\mathbf{if}$
statement is $\Integer \cup \String$, so the assignment looks fine.
However, the projected type of $r$ inside the $\mathbf{if}$ branch is $\Integer$,
not matching the $\String$ value that $r$ has after the assignment.
